---
title: "EM Algorithm - Survival Data"
author: "Matthew Scott | mscott24@bu.edu"
date: "January 2023"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<br>

**Question** Let $t_{1}, t_{2}, ..., t_{n}$ i.d.d. with pdf $\rho exp(-\rho t)$, i.e., exponential with parameter $\rho$. Supposed that we observe $(y_{i}, \delta_{i})$, where $y_{i}=min(t_{i}, c_{i})$ and $\delta_{i}=i$ if $t_{i}<c_{i}$ and 0 otherwise for $i=1,2,...,n$. Implement an EM-algorithm for the Stanford Heart Transplant data (for the variables survival time $y$ and survival status $\delta$) and compute the standard error of the estimate $\hat{\rho}_{MLE}$.

<br>

**Derivation** 
Assume that $Y_{1},...,Y_{m}$ are non-censored data and the rest of the observation are censored with censoring times given by $c_{m+1}, ..., c_{n}$. Let $Z_{m+1},..., Z_{n}$ be the survival times for the censored data, with the complete data likelihood given by

$$
\begin{aligned}
  p(\rho | y, z) & \propto \rho^{n} exp[-\rho (\sum_{i=1}^{m}y_{i}+\sum_{i=m+1}^{n}z_{i})] \\
  log p(\rho | y, z) & \propto n log \rho -\rho (\sum_{i=1}^{m}y_{i}+\sum_{i=m+1}^{n}z_{i}) \\
\end{aligned}
$$

The conditional predictive distribution of $z$ given $Z>c$ is a truncated exponential distribution. Due to the memoryless property of exponential distributions, we have
$$
\begin{aligned} 
E(Z_{i}|Z_{i}>c_{i}, \rho^{(k)}) & = c_{i}+\frac{1}{\rho^{(k)}} \\ 
\end{aligned}
$$
Thus
$$
\begin{aligned} 
  Q(\rho, \rho^{(k)}) & = n log \rho - \rho (\sum_{i=1}^{m} y_{i} + \sum_{i=m+1}^{n} (c_{i} + \frac{1}{\rho^{(k)}}))\\
\end{aligned}
$$

In the M-step, we maximize $Q(\rho, \rho^{(k)})$, leading to the EM update
$$
\begin{aligned} 
  \rho^{(k+1)} & = \frac{n} {\sum_{i=1}^{m} y_{i} + \sum_{i=m+1}^{n} (c_{i} + \frac{1}{\rho^{(k)}})} \\
\end{aligned}
$$

The SE can be found via Louis's method
$$
\begin{aligned} 
      \frac{d^{2}}{d \rho^{2}}  Q(\rho, \rho_{EM}) & =  \frac{n}{\rho_{EM}^{2}} \\
      \rightarrow Var(\frac{d}{d \rho} log p(\rho | y, z)) & = \sum_{i=m+1}^{n}  Var(z_{i}|z_{i} > c_{i}, \rho_{EM}) = \frac{n-m}{\rho_{EM}^{2}} \\
\end{aligned}
$$

Thus the observed Fisher information evaulated at $\rho_{EM}$ is 
$$
\begin{aligned} 
  \frac{n}{\rho_{EM}^{2}}-\frac{n-m}{\rho_{EM}^{2}} &= \frac{m}{\rho_{EM}^{2}} \\
  \rightarrow \boxed{SE=\frac{\hat{\rho}_{EM}}{\sqrt(m)}} \\
\end{aligned}
$$
<br>

**EM Algorithm**
```{r}
library(survival)

#initialize
r <- rep(NA,1000)
r[1] <- 0.0001
n <- length(stanford2$time)

#perform iterations
for(k in 1:999) {
  r[k+1]<-n/{sum(stanford2$time[stanford2$status==0]) + sum(stanford2$time[stanford2$status==1]+(1/r[k]))}
  mle_em <- r[k+1]
  #stop is convergence is reached
  if(abs(r[k+1]-r[k]) < 10^(-10)) {break}
}
#calculate SE
se_em <- mle_em/sqrt(length(stanford2$time[stanford2$status==0]))
```
<br>

**Estimated $\hat{\rho}_{MLE}$**
```{r}
mle_em
```
<br>

**SE of $\hat{\rho}_{MLE}$**
```{r}
se_em
```